---
title: "Album Cover Similarity Analysis"
subtitle: "Interactive visual comparison tool for Taylor Swift and Red Hot Chili Peppers"
format:
  html:
    toc: true
    code-fold: true
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(jsonlite)

albums <- read_csv("album_metadata.csv", show_col_types = FALSE)
pairs <- read_csv("album_similarity_pairs.csv", show_col_types = FALSE)

# Build symmetrical pairwise relationships
pairs_sym <- bind_rows(
  pairs |> transmute(target = title1, other_title = title2, other_artist = artist2, other_img = img2, similarity, color_sim, design_sim),
  pairs |> transmute(target = title2, other_title = title1, other_artist = artist1, other_img = img1, similarity, color_sim, design_sim)
)

album_data <- list()
for (i in seq_len(nrow(albums))) {
  t <- albums$clean_title[i]
  rel <- pairs_sym |> filter(target == t)
  album_data[[t]] <- list(
    title = t,
    artist = albums$artist[i],
    img = albums$image_path[i],
    overall_most = rel |> arrange(desc(similarity)) |> slice_head(n = 3),
    overall_least = rel |> arrange(similarity) |> slice_head(n = 3),
    color_most = rel |> arrange(desc(color_sim)) |> slice_head(n = 3),
    color_least = rel |> arrange(color_sim) |> slice_head(n = 3),
    design_most = rel |> arrange(desc(design_sim)) |> slice_head(n = 3),
    design_least = rel |> arrange(design_sim) |> slice_head(n = 3)
  )
}

json_payload <- toJSON(album_data, auto_unbox = TRUE)
```

## How Similarity Is Measured

Album covers are compared across two complementary visual dimensions:

1. **Color (HSV Palette):** Compares distribution of hue, color saturation, and brightness.
2. **Design (Spatial Composition):** Compares spatial layout, lighting balance, and structural shapes.

Scores range from **-1.0** (complete opposites) to **+1.0** (virtually identical).

::: {.callout-tip}
### Explore 1,000 Pop Albums
Looking for more? Browse the [**1,000 Pop Albums Catalog**](pop-albums.qmd) with interactive search, filters, and cover art thumbnails.
:::

---

## Interactive Album Comparison Tool

Pick any album below to instantly inspect its **most similar** and **least similar** matches.

```{r}
#| label: interactive-ui
#| results: asis
#| echo: false

# Safely output JSON
cat('<script type="application/json" id="album-similarity-data">')
cat(json_payload)
cat('</script>\n')

# Render controls
cat('
<div style="background: #f8fafc; border: 1px solid #e2e8f0; border-radius: 10px; padding: 18px; margin-bottom: 24px;">
  <div style="display: flex; gap: 20px; flex-wrap: wrap; align-items: flex-end;">
    <div style="flex: 1; min-width: 260px;">
      <label for="album-picker" style="font-weight: 600; display: block; margin-bottom: 6px; color: #1e293b;">
        Select an Album:
      </label>
      <select id="album-picker" style="width: 100%; padding: 8px 12px; border-radius: 6px; border: 1px solid #cbd5e1; font-size: 1rem;">
')

ts_albums <- albums |> filter(artist == "Taylor Swift")
cat('        <optgroup label="Taylor Swift">\n')
for (i in seq_len(nrow(ts_albums))) {
  sel <- if (ts_albums$clean_title[i] == "folklore") ' selected' else ''
  cat(sprintf('          <option value="%s"%s>%s</option>\n', ts_albums$clean_title[i], sel, ts_albums$clean_title[i]))
}
cat('        </optgroup>\n')

rhcp_albums <- albums |> filter(artist == "Red Hot Chili Peppers")
cat('        <optgroup label="Red Hot Chili Peppers">\n')
for (i in seq_len(nrow(rhcp_albums))) {
  cat(sprintf('          <option value="%s">%s</option>\n', rhcp_albums$clean_title[i], rhcp_albums$clean_title[i]))
}
cat('        </optgroup>\n')

cat('
      </select>
    </div>
    <div style="width: 240px;">
      <label for="metric-picker" style="font-weight: 600; display: block; margin-bottom: 6px; color: #1e293b;">
        Compare By:
      </label>
      <select id="metric-picker" style="width: 100%; padding: 8px 12px; border-radius: 6px; border: 1px solid #cbd5e1; font-size: 1rem;">
        <option value="overall" selected>Combined (Color & Design)</option>
        <option value="color">Color Palette Only</option>
        <option value="design">Spatial Design Only</option>
      </select>
    </div>
  </div>
</div>

<div id="comparison-display" style="display: grid; grid-template-columns: 280px 1fr; gap: 24px; align-items: start;">
  <!-- Left column: Chosen album -->
  <div id="selected-card" style="background: #ffffff; border: 2px solid #3b82f6; border-radius: 10px; padding: 18px; text-align: center; box-shadow: 0 4px 6px -1px rgba(59, 130, 246, 0.1);">
    <!-- Injected by JS -->
  </div>

  <!-- Right column: Most & Least Similar lists -->
  <div style="display: flex; flex-direction: column; gap: 20px;">
    <!-- Most Similar -->
    <div style="background: #f0fdf4; border: 1px solid #bbf7d0; border-radius: 10px; padding: 16px;">
      <h4 style="margin-top: 0; margin-bottom: 12px; color: #166534; display: flex; align-items: center; gap: 8px;">
        <span>🟢</span> Most Similar Album Covers
      </h4>
      <div id="most-similar-cards" style="display: grid; grid-template-columns: repeat(auto-fill, minmax(180px, 1fr)); gap: 14px;">
        <!-- Injected by JS -->
      </div>
    </div>

    <!-- Least Similar -->
    <div style="background: #fef2f2; border: 1px solid #fecaca; border-radius: 10px; padding: 16px;">
      <h4 style="margin-top: 0; margin-bottom: 12px; color: #991b1b; display: flex; align-items: center; gap: 8px;">
        <span>🔴</span> Least Similar Album Covers
      </h4>
      <div id="least-similar-cards" style="display: grid; grid-template-columns: repeat(auto-fill, minmax(180px, 1fr)); gap: 14px;">
        <!-- Injected by JS -->
      </div>
    </div>
  </div>
</div>

<script>
document.addEventListener("DOMContentLoaded", function() {
  const dataElement = document.getElementById("album-similarity-data");
  if (!dataElement) return;
  const albumData = JSON.parse(dataElement.textContent);

  const albumPicker = document.getElementById("album-picker");
  const metricPicker = document.getElementById("metric-picker");
  const selectedCard = document.getElementById("selected-card");
  const mostCards = document.getElementById("most-similar-cards");
  const leastCards = document.getElementById("least-similar-cards");

  function renderCard(item, metric, isMatch) {
    let score = "";
    if (metric === "color") {
      score = `Color: ${(item.color_sim * 100).toFixed(1)}%`;
    } else if (metric === "design") {
      score = `Design: ${(item.design_sim * 100).toFixed(1)}%`;
    } else {
      score = `Match: ${(item.similarity * 100).toFixed(1)}%`;
    }

    const badgeBg = isMatch ? "#dcfce7" : "#fee2e2";
    const badgeColor = isMatch ? "#15803d" : "#b91c1c";

    return `
      <div style="background: #ffffff; border: 1px solid #e2e8f0; border-radius: 8px; padding: 10px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.05);">
        <img src="${item.other_img}" style="width: 100%; aspect-ratio: 1/1; object-fit: cover; border-radius: 4px; border: 1px solid #e2e8f0;" alt="${item.other_title}"/>
        <div style="font-weight: 600; font-size: 0.85rem; margin-top: 8px; overflow: hidden; text-overflow: ellipsis; white-space: nowrap;" title="${item.other_title}">${item.other_title}</div>
        <div style="font-size: 0.75rem; color: #64748b; margin-bottom: 6px;">${item.other_artist}</div>
        <div style="display: inline-block; background: ${badgeBg}; color: ${badgeColor}; font-weight: 600; font-size: 0.75rem; padding: 2px 8px; border-radius: 12px;">
          ${score}
        </div>
        <div style="font-size: 0.68rem; color: #94a3b8; margin-top: 4px;">
          Color: ${(item.color_sim * 100).toFixed(0)}% | Design: ${(item.design_sim * 100).toFixed(0)}%
        </div>
      </div>
    `;
  }

  function updateDisplay() {
    const key = albumPicker.value;
    const metric = metricPicker.value;
    const current = albumData[key];
    if (!current) return;

    // Render left selected card
    selectedCard.innerHTML = `
      <div style="font-size: 0.75rem; text-transform: uppercase; font-weight: 700; color: #3b82f6; letter-spacing: 0.05em; margin-bottom: 6px;">Selected Album</div>
      <img src="${current.img}" style="width: 100%; aspect-ratio: 1/1; object-fit: cover; border-radius: 6px; border: 1px solid #cbd5e1;" alt="${current.title}"/>
      <h3 style="margin: 12px 0 4px 0; font-size: 1.15rem; color: #0f172a;">${current.title}</h3>
      <div style="font-size: 0.9rem; color: #64748b; font-weight: 500;">${current.artist}</div>
      <div style="margin-top: 10px; font-size: 0.75rem; color: #475569; background: #eff6ff; padding: 6px; border-radius: 6px;">
        Comparing against all other albums by <strong>${metric === "color" ? "Color Palette" : metric === "design" ? "Spatial Design" : "Color & Design"}</strong>.
      </div>
    `;

    // Pick top and least lists based on metric
    const mostList = current[metric + "_most"] || current.overall_most;
    const leastList = current[metric + "_least"] || current.overall_least;

    mostCards.innerHTML = mostList.map(item => renderCard(item, metric, true)).join("");
    leastCards.innerHTML = leastList.map(item => renderCard(item, metric, false)).join("");
  }

  albumPicker.addEventListener("change", updateDisplay);
  metricPicker.addEventListener("change", updateDisplay);

  updateDisplay();
});
</script>
')
```

---

## All Analyzed Albums

```{r}
#| label: table
#| echo: false

albums |>
  select(Artist = artist, Album = clean_title, `Release Date` = release_date) |>
  knitr::kable()
```
