---
title: "Album Cover Similarity Analysis"
subtitle: "Interactive visual comparison tool across 1,000+ pop and rock albums"
format:
  html:
    toc: true
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(jsonlite)
library(DT)

pop_albums <- read_csv("pop_albums.csv", show_col_types = FALSE)

# Build compact feature dataset for instant client-side similarity comparison
album_catalog <- pop_albums |>
  mutate(
    hex_clean = sub("^#", "", hex_color),
    r = strtoi(substr(hex_clean, 1, 2), 16L) / 255,
    g = strtoi(substr(hex_clean, 3, 4), 16L) / 255,
    b_val = strtoi(substr(hex_clean, 5, 6), 16L) / 255,
    b_norm = brightness / 100,
    s_norm = saturation / 100,
    c_norm = contrast / 100
  ) |>
  transmute(
    id = row_number(),
    artist,
    album,
    year,
    cover = cover_full,
    thumb = cover_thumbnail,
    hex = hex_color,
    palette,
    r = round(r, 3),
    g = round(g, 3),
    b = round(b_val, 3),
    bright = round(b_norm, 3),
    sat = round(s_norm, 3),
    contrast = round(c_norm, 3)
  )

json_payload <- toJSON(album_catalog, auto_unbox = TRUE)
```

## How Similarity Is Measured

Every album cover is compared across two visual dimensions:

1. **Color Palette:** Compares dominant RGB hues, saturation (vibrancy), and color tone.
2. **Design & Composition:** Compares overall brightness balance and grayscale visual contrast.

Scores range from **0%** (completely opposite visual styles) to **100%** (virtually identical color & layout).

---

## Interactive Album Comparison Tool

Pick any artist and album below to instantly compare its cover against the entire catalog of **1,000+ albums**.

```{r}
#| label: json-data
#| results: asis
#| echo: false

cat('<script type="application/json" id="album-similarity-data">\n')
cat(json_payload)
cat('\n</script>\n')
```

```{=html}
<div style="background: #f8fafc; border: 1px solid #e2e8f0; border-radius: 10px; padding: 18px; margin-bottom: 24px;">
  <div style="display: flex; gap: 16px; flex-wrap: wrap; align-items: flex-end;">
    <div style="flex: 1; min-width: 220px;">
      <label for="artist-picker" style="font-weight: 600; display: block; margin-bottom: 6px; color: #1e293b;">
        1. Pick an Artist:
      </label>
      <select id="artist-picker" style="width: 100%; padding: 8px 12px; border-radius: 6px; border: 1px solid #cbd5e1; font-size: 1rem; background: white;">
      </select>
    </div>

    <div style="flex: 1.5; min-width: 260px;">
      <label for="album-picker" style="font-weight: 600; display: block; margin-bottom: 6px; color: #1e293b;">
        2. Pick an Album:
      </label>
      <select id="album-picker" style="width: 100%; padding: 8px 12px; border-radius: 6px; border: 1px solid #cbd5e1; font-size: 1rem; background: white;">
      </select>
    </div>

    <div style="width: 220px;">
      <label for="metric-picker" style="font-weight: 600; display: block; margin-bottom: 6px; color: #1e293b;">
        3. Compare By:
      </label>
      <select id="metric-picker" style="width: 100%; padding: 8px 12px; border-radius: 6px; border: 1px solid #cbd5e1; font-size: 1rem; background: white;">
        <option value="overall" selected>Combined (Color & Design)</option>
        <option value="color">Color Palette Only</option>
        <option value="design">Spatial Design Only</option>
      </select>
    </div>
  </div>
</div>

<div id="comparison-display" style="display: grid; grid-template-columns: 280px 1fr; gap: 24px; align-items: start;">
  <div id="selected-card" style="background: #ffffff; border: 2px solid #3b82f6; border-radius: 10px; padding: 18px; text-align: center; box-shadow: 0 4px 6px -1px rgba(59, 130, 246, 0.1);">
  </div>

  <div style="display: flex; flex-direction: column; gap: 20px;">
    <div style="background: #f0fdf4; border: 1px solid #bbf7d0; border-radius: 10px; padding: 16px;">
      <h4 style="margin-top: 0; margin-bottom: 12px; color: #166534; display: flex; align-items: center; gap: 8px;">
        <span>🟢</span> Most Similar Album Covers Across the Catalog
      </h4>
      <div id="most-similar-cards" style="display: grid; grid-template-columns: repeat(auto-fill, minmax(180px, 1fr)); gap: 14px;">
      </div>
    </div>

    <div style="background: #fef2f2; border: 1px solid #fecaca; border-radius: 10px; padding: 16px;">
      <h4 style="margin-top: 0; margin-bottom: 12px; color: #991b1b; display: flex; align-items: center; gap: 8px;">
        <span>🔴</span> Least Similar Album Covers Across the Catalog
      </h4>
      <div id="least-similar-cards" style="display: grid; grid-template-columns: repeat(auto-fill, minmax(180px, 1fr)); gap: 14px;">
      </div>
    </div>
  </div>
</div>

<script>
function initAlbumComparison() {
  const dataElement = document.getElementById("album-similarity-data");
  if (!dataElement) return;
  const albumData = JSON.parse(dataElement.textContent);

  const artistPicker = document.getElementById("artist-picker");
  const albumPicker = document.getElementById("album-picker");
  const metricPicker = document.getElementById("metric-picker");
  const selectedCard = document.getElementById("selected-card");
  const mostCards = document.getElementById("most-similar-cards");
  const leastCards = document.getElementById("least-similar-cards");

  if (!artistPicker || !albumPicker || !metricPicker || !selectedCard || !mostCards || !leastCards) return;

  const allArtists = Array.from(new Set(albumData.map(a => a.artist))).sort((a, b) => a.localeCompare(b));
  const featured = ["Taylor Swift", "Red Hot Chili Peppers", "Billie Eilish", "Harry Styles", "Bruno Mars", "Olivia Rodrigo", "Dua Lipa", "Ariana Grande"];
  
  artistPicker.innerHTML = "";
  
  const featuredGroup = document.createElement("optgroup");
  featuredGroup.label = "Featured Artists";
  featured.forEach(name => {
    if (allArtists.includes(name)) {
      const opt = document.createElement("option");
      opt.value = name;
      opt.textContent = name;
      if (name === "Taylor Swift") opt.selected = true;
      featuredGroup.appendChild(opt);
    }
  });
  artistPicker.appendChild(featuredGroup);

  const allGroup = document.createElement("optgroup");
  allGroup.label = "All Artists (A-Z)";
  allArtists.forEach(name => {
    const opt = document.createElement("option");
    opt.value = name;
    opt.textContent = name;
    allGroup.appendChild(opt);
  });
  artistPicker.appendChild(allGroup);

  function populateAlbumsForArtist(artistName, defaultAlbum) {
    albumPicker.innerHTML = "";
    const list = albumData.filter(a => a.artist === artistName);
    list.forEach((item, idx) => {
      const opt = document.createElement("option");
      opt.value = item.album;
      opt.textContent = item.album;
      if (defaultAlbum ? item.album === defaultAlbum : idx === 0) {
        opt.selected = true;
      }
      albumPicker.appendChild(opt);
    });
  }

  function renderCard(item, metric, isMatch) {
    let scoreText = "";
    if (metric === "color") {
      scoreText = `Color: ${(item.colorSim * 100).toFixed(1)}%`;
    } else if (metric === "design") {
      scoreText = `Design: ${(item.designSim * 100).toFixed(1)}%`;
    } else {
      scoreText = `Match: ${(item.overallSim * 100).toFixed(1)}%`;
    }

    const badgeBg = isMatch ? "#dcfce7" : "#fee2e2";
    const badgeColor = isMatch ? "#15803d" : "#b91c1c";

    return `
      <div style="background: #ffffff; border: 1px solid #e2e8f0; border-radius: 8px; padding: 10px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.05);">
        <img src="${item.cover}" onerror="this.onerror=null;this.src='${item.thumb}'" style="width: 100%; aspect-ratio: 1/1; object-fit: cover; border-radius: 4px; border: 1px solid #e2e8f0; display: block;" alt="${item.album}"/>
        <div style="font-weight: 600; font-size: 0.85rem; margin-top: 8px; overflow: hidden; text-overflow: ellipsis; white-space: nowrap;" title="${item.album}">${item.album}</div>
        <div style="font-size: 0.75rem; color: #64748b; margin-bottom: 6px; overflow: hidden; text-overflow: ellipsis; white-space: nowrap;">${item.artist} (${item.year})</div>
        <div style="display: inline-block; background: ${badgeBg}; color: ${badgeColor}; font-weight: 600; font-size: 0.75rem; padding: 2px 8px; border-radius: 12px;">
          ${scoreText}
        </div>
        <div style="font-size: 0.68rem; color: #94a3b8; margin-top: 4px;">
          Color: ${(item.colorSim * 100).toFixed(0)}% | Design: ${(item.designSim * 100).toFixed(0)}%
        </div>
      </div>
    `;
  }

  function updateDisplay() {
    const artistName = artistPicker.value;
    const albumName = albumPicker.value;
    const metric = metricPicker.value;
    
    const target = albumData.find(a => a.artist === artistName && a.album === albumName);
    if (!target) return;

    selectedCard.innerHTML = `
      <div style="font-size: 0.75rem; text-transform: uppercase; font-weight: 700; color: #3b82f6; letter-spacing: 0.05em; margin-bottom: 6px;">Selected Album</div>
      <img src="${target.cover}" onerror="this.onerror=null;this.src='${target.thumb}'" style="width: 100%; aspect-ratio: 1/1; object-fit: cover; border-radius: 6px; border: 1px solid #cbd5e1; display: block;" alt="${target.album}"/>
      <h3 style="margin: 12px 0 4px 0; font-size: 1.15rem; color: #0f172a;">${target.album}</h3>
      <div style="font-size: 0.9rem; color: #64748b; font-weight: 500;">${target.artist} &bull; ${target.year}</div>
      <div style="margin-top: 10px; font-size: 0.75rem; color: #475569; background: #eff6ff; padding: 8px; border-radius: 6px; text-align: left;">
        <div><strong>Palette:</strong> <span style="display:inline-block;width:12px;height:12px;border-radius:2px;background:${target.hex};vertical-align:middle;border:1px solid #cbd5e1;"></span> ${target.palette}</div>
        <div><strong>Brightness:</strong> ${(target.bright * 100).toFixed(0)}% | <strong>Vibrancy:</strong> ${(target.sat * 100).toFixed(0)}%</div>
      </div>
    `;

    const candidates = [];
    for (let i = 0; i < albumData.length; i++) {
      const a = albumData[i];
      if (a.id === target.id) continue;

      const dColor = Math.sqrt(
        Math.pow(a.r - target.r, 2) +
        Math.pow(a.g - target.g, 2) +
        Math.pow(a.b - target.b, 2) +
        Math.pow(a.sat - target.sat, 2)
      ) / 2.0;
      const colorSim = Math.max(0, 1 - dColor);

      const dDesign = Math.sqrt(
        Math.pow(a.bright - target.bright, 2) +
        Math.pow(a.contrast - target.contrast, 2)
      ) / 1.414;
      const designSim = Math.max(0, 1 - dDesign);

      const overallSim = 0.5 * colorSim + 0.5 * designSim;

      candidates.push({
        ...a,
        colorSim,
        designSim,
        overallSim
      });
    }

    if (metric === "color") {
      candidates.sort((x, y) => y.colorSim - x.colorSim);
    } else if (metric === "design") {
      candidates.sort((x, y) => y.designSim - x.designSim);
    } else {
      candidates.sort((x, y) => y.overallSim - x.overallSim);
    }

    const mostList = candidates.slice(0, 3);
    const leastList = candidates.slice(-3).reverse();

    mostCards.innerHTML = mostList.map(item => renderCard(item, metric, true)).join("");
    leastCards.innerHTML = leastList.map(item => renderCard(item, metric, false)).join("");
  }

  artistPicker.addEventListener("change", function() {
    populateAlbumsForArtist(artistPicker.value);
    updateDisplay();
  });

  albumPicker.addEventListener("change", updateDisplay);
  metricPicker.addEventListener("change", updateDisplay);

  populateAlbumsForArtist("Taylor Swift", "Lover");
  updateDisplay();
}

if (document.readyState === "loading") {
  document.addEventListener("DOMContentLoaded", initAlbumComparison);
} else {
  initAlbumComparison();
}
</script>
```

---

## All Analyzed Albums (1,000+ Pop & Rock Collection)

Browse and search the entire collection of album covers below:

```{r}
#| label: table
#| echo: false

pop_albums |>
  mutate(
    Cover = sprintf('<img src="%s" width="50" height="50" style="border-radius: 4px; object-fit: cover;" alt="%s" loading="lazy"/>', cover_thumbnail, album),
    Color = sprintf('<span style="display:inline-block; width:20px; height:20px; background-color:%s; border:1px solid #cbd5e1; border-radius:3px; vertical-align:middle;" title="%s"></span>', hex_color, hex_color)
  ) |>
  select(Cover, Color, Artist = artist, Album = album, Year = year, Genre = genre, Palette = palette) |>
  datatable(
    options = list(pageLength = 10, searchHighlight = TRUE),
    escape = FALSE,
    rownames = FALSE
  )
```
