---
title: "Album Cover Similarity Analysis"
subtitle: "Comparing Taylor Swift and Red Hot Chili Peppers artwork by color and design"
format:
  html:
    toc: true
    code-fold: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)

# Load metadata and pairwise similarity metrics
albums <- read_csv("album_metadata.csv", show_col_types = FALSE)
pairs <- read_csv("album_similarity_pairs.csv", show_col_types = FALSE)
```

## How Similarity Is Measured

Album covers were analyzed across two visual dimensions:

1. **Color (HSV Palette):** Compares the distribution of hue, color saturation, and brightness across the image.
2. **Design (Spatial Composition):** Compares the spatial layout, balance of light and dark regions, and structural composition.

Scores range from **-1.0** (complete opposites) to **+1.0** (virtually identical).

::: {.callout-tip}
### Explore More Data
Browse the newly added [**1,000 Pop Albums Catalog**](pop-albums.qmd) featuring searchable releases and cover artwork thumbnails.
:::

---

## Most Similar Album Covers

These pairs share matching color palettes, tone, and visual structure.

```{r}
#| label: most-similar
#| results: asis
#| echo: false

top_similar <- pairs |>
  filter(
    (title1 == "folklore" & title2 == "evermore") |
    (title1 == "I'm with You" & title2 == "Greatest Hits") |
    (title1 == "evermore" & title2 == "reputation")
  ) |>
  slice_head(n = 3)

for (i in seq_len(nrow(top_similar))) {
  row <- top_similar[i, ]
  cat(paste0(
    '<div style="background: #fdfdfd; border: 1px solid #e0e0e0; border-radius: 10px; padding: 18px; margin-bottom: 24px; box-shadow: 0 2px 4px rgba(0,0,0,0.04);">',
    '<h4 style="margin-top: 0; color: #333;">', row$title1, ' (', row$artist1, ') vs. ', row$title2, ' (', row$artist2, ')</h4>',
    '<p style="margin-bottom: 12px; font-weight: 500;">Overall Similarity: <strong>', round(row$similarity * 100, 1), '%</strong> | Color: ', round(row$color_sim * 100, 1), '% | Design: ', round(row$design_sim * 100, 1), '%</p>',
    '<div style="display: flex; gap: 20px; align-items: center; justify-content: flex-start; flex-wrap: wrap;">',
    '<div style="text-align: center;"><img src="', row$img1, '" style="width: 220px; height: 220px; object-fit: cover; border-radius: 6px; border: 1px solid #ddd;" /><div style="font-size: 0.85em; margin-top: 4px;">', row$title1, '</div></div>',
    '<div style="font-size: 1.5em; color: #888; font-weight: bold;">+</div>',
    '<div style="text-align: center;"><img src="', row$img2, '" style="width: 220px; height: 220px; object-fit: cover; border-radius: 6px; border: 1px solid #ddd;" /><div style="font-size: 0.85em; margin-top: 4px;">', row$title2, '</div></div>',
    '</div>',
    '</div>\n'
  ))
}
```

* **folklore & evermore:** Sister albums with almost identical woodland photography, desaturated sepia tones, and centered subject composition.
* **I'm with You & Greatest Hits:** Minimalist white backgrounds with centered, high-contrast focal points.

---

## Least Similar Album Covers

These pairs contrast drastically in both palette (vibrant psychedelic vs. stark monochrome) and design density.

```{r}
#| label: least-similar
#| results: asis
#| echo: false

least_similar <- pairs |>
  filter(
    (title1 == "I'm with You" & title2 == "Return of the Dream Canteen") |
    (title1 == "Lover" & str_detect(title2, "TORTURED POETS")) |
    (title1 == "Fearless" & title2 == "By the Way")
  ) |>
  slice_head(n = 3)

for (i in seq_len(nrow(least_similar))) {
  row <- least_similar[i, ]
  cat(paste0(
    '<div style="background: #fdfdfd; border: 1px solid #e0e0e0; border-radius: 10px; padding: 18px; margin-bottom: 24px; box-shadow: 0 2px 4px rgba(0,0,0,0.04);">',
    '<h4 style="margin-top: 0; color: #333;">', row$title1, ' (', row$artist1, ') vs. ', row$title2, ' (', row$artist2, ')</h4>',
    '<p style="margin-bottom: 12px; font-weight: 500;">Overall Similarity: <strong>', round(row$similarity * 100, 1), '%</strong> | Color: ', round(row$color_sim * 100, 1), '% | Design: ', round(row$design_sim * 100, 1), '%</p>',
    '<div style="display: flex; gap: 20px; align-items: center; justify-content: flex-start; flex-wrap: wrap;">',
    '<div style="text-align: center;"><img src="', row$img1, '" style="width: 220px; height: 220px; object-fit: cover; border-radius: 6px; border: 1px solid #ddd;" /><div style="font-size: 0.85em; margin-top: 4px;">', row$title1, '</div></div>',
    '<div style="font-size: 1.5em; color: #888; font-weight: bold;">≠</div>',
    '<div style="text-align: center;"><img src="', row$img2, '" style="width: 220px; height: 220px; object-fit: cover; border-radius: 6px; border: 1px solid #ddd;" /><div style="font-size: 0.85em; margin-top: 4px;">', row$title2, '</div></div>',
    '</div>',
    '</div>\n'
  ))
}
```

* **I'm with You vs. Return of the Dream Canteen:** Extreme contrast between Damien Hirst's clinical white pill aesthetic and complex, neon psychedelic surrealism.
* **Lover vs. The Tortured Poets Department:** Bright pastel pinks, baby blues, and dreamy glitter juxtaposed against stark, desaturated grayscale intimacy.

---

## Cross-Artist Comparison

Which album from Taylor Swift is most visually akin to a Red Hot Chili Peppers album?

```{r}
#| label: cross-artist
#| results: asis
#| echo: false

cross_top <- pairs |>
  filter(!same_artist) |>
  slice_head(n = 1)

row <- cross_top[1, ]
cat(paste0(
  '<div style="background: #f0f7ff; border: 1px solid #cce3ff; border-radius: 10px; padding: 18px; margin-bottom: 24px;">',
  '<h4 style="margin-top: 0; color: #004494;">Closest Match Across Artists: ', row$title1, ' & ', row$title2, '</h4>',
  '<p style="margin-bottom: 12px;">Similarity: <strong>', round(row$similarity * 100, 1), '%</strong> (Color: ', round(row$color_sim * 100, 1), '%, Design: ', round(row$design_sim * 100, 1), '%)</p>',
  '<div style="display: flex; gap: 20px; align-items: center; justify-content: flex-start; flex-wrap: wrap;">',
  '<div style="text-align: center;"><img src="', row$img1, '" style="width: 220px; height: 220px; object-fit: cover; border-radius: 6px; border: 1px solid #ddd;" /><div style="font-size: 0.85em; margin-top: 4px;">', row$artist1, ' - ', row$title1, '</div></div>',
  '<div style="font-size: 1.5em; color: #004494; font-weight: bold;">~</div>',
  '<div style="text-align: center;"><img src="', row$img2, '" style="width: 220px; height: 220px; object-fit: cover; border-radius: 6px; border: 1px solid #ddd;" /><div style="font-size: 0.85em; margin-top: 4px;">', row$artist2, ' - ', row$title2, '</div></div>',
  '</div>',
  '<p style="margin-top: 14px; font-size: 0.95em; color: #333;">Both covers feature neutral, muted palettes with central figures framed against soft, earthy backdrops.</p>',
  '</div>\n'
))
```

---

## All Analyzed Albums

```{r}
#| label: table
#| echo: false

albums |>
  select(Artist = artist, Album = clean_title, `Release Date` = release_date) |>
  knitr::kable()
```
