---
title: "1,000 Pop Album Covers: Visual Comparison"
subtitle: "Explore how 1,000 pop album covers compare in color palette, brightness, and vibrancy"
format:
  html:
    page-layout: full
    toc: true
    code-fold: true
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(plotly)
library(DT)

pop_albums <- read_csv("pop_albums.csv", show_col_types = FALSE)
```

Understanding and comparing 1,000 album covers at once can be overwhelming. To make comparisons intuitive, every album cover was analyzed along two key visual axes:

- **Brightness (0–100%):** From shadow-heavy, pitch-black photography to stark white minimalist backgrounds.
- **Color Saturation / Vibrancy (0–100%):** From black-and-white monochrome to neon, hyper-saturated palettes.

---

## 1. Visual Comparison Map

Hover over any point below to see the album title, artist, year, and its visual profile.

```{r}
#| label: scatter-plot
#| echo: false

p <- pop_albums |>
  ggplot(aes(
    x = brightness,
    y = saturation,
    color = palette,
    text = paste0(
      "<b>", htmltools::htmlEscape(album), "</b><br>",
      "Artist: ", htmltools::htmlEscape(artist), "<br>",
      "Year: ", year, "<br>",
      "Palette: ", palette, "<br>",
      "Brightness: ", brightness, "%<br>",
      "Vibrancy: ", saturation, "%"
    )
  )) +
  geom_point(alpha = 0.65, size = 2.2) +
  scale_color_manual(values = c(
    "Cool Blue" = "#2563eb",
    "Dark & Moody" = "#1e293b",
    "Green" = "#16a34a",
    "Minimalist White" = "#94a3b8",
    "Monochrome / Gray" = "#64748b",
    "Pink & Magenta" = "#db2777",
    "Warm Red & Orange" = "#ea580c",
    "Warm Yellow & Gold" = "#ca8a04"
  )) +
  labs(
    title = "Color Saturation vs. Brightness across 1,000 Pop Covers",
    x = "Brightness (% 0 = Dark, 100 = Light)",
    y = "Color Vibrancy (% 0 = Grayscale, 100 = Vivid)",
    color = "Palette Style"
  ) +
  theme_minimal(base_size = 11)

ggplotly(p, tooltip = "text")
```

---

## 2. Extreme Comparisons: The Visual Spectrum

Here is what the extremes of the 1,000 album collection look like when placed side-by-side.

### Brightest vs. Darkest Covers

```{r}
#| label: light-vs-dark
#| results: asis
#| echo: false

bright_albums <- pop_albums |> arrange(desc(brightness)) |> slice_head(n = 3)
dark_albums <- pop_albums |> arrange(brightness) |> slice_head(n = 3)

cat('<div style="display: grid; grid-template-columns: 1fr 1fr; gap: 24px; margin-bottom: 24px;">\n')

# Brightest column
cat('<div style="background: #f8fafc; border: 1px solid #e2e8f0; border-radius: 8px; padding: 16px;">\n')
cat('<h4 style="margin-top:0; color: #0284c7;">☀️ Brightest (High Exposure & White)</h4>\n')
cat('<div style="display: flex; gap: 12px; flex-wrap: wrap;">\n')
for (i in seq_len(nrow(bright_albums))) {
  cat(paste0(
    '<div style="text-align: center; width: 110px;">',
    '<img src="', bright_albums$cover_thumbnail[i], '" style="width: 100px; height: 100px; object-fit: cover; border-radius: 6px; border: 1px solid #ccc;"/>',
    '<div style="font-size: 0.75em; font-weight: bold; margin-top: 4px; overflow: hidden; text-overflow: ellipsis; white-space: nowrap;">', htmltools::htmlEscape(bright_albums$album[i]), '</div>',
    '<div style="font-size: 0.7em; color: #64748b;">', bright_albums$brightness[i], '% Light</div>',
    '</div>'
  ))
}
cat('</div></div>\n')

# Darkest column
cat('<div style="background: #0f172a; color: #f8fafc; border-radius: 8px; padding: 16px;">\n')
cat('<h4 style="margin-top:0; color: #38bdf8;">🌑 Darkest (Shadows & Low-Key)</h4>\n')
cat('<div style="display: flex; gap: 12px; flex-wrap: wrap;">\n')
for (i in seq_len(nrow(dark_albums))) {
  cat(paste0(
    '<div style="text-align: center; width: 110px;">',
    '<img src="', dark_albums$cover_thumbnail[i], '" style="width: 100px; height: 100px; object-fit: cover; border-radius: 6px; border: 1px solid #334155;"/>',
    '<div style="font-size: 0.75em; font-weight: bold; margin-top: 4px; color: #e2e8f0; overflow: hidden; text-overflow: ellipsis; white-space: nowrap;">', htmltools::htmlEscape(dark_albums$album[i]), '</div>',
    '<div style="font-size: 0.7em; color: #94a3b8;">', dark_albums$brightness[i], '% Light</div>',
    '</div>'
  ))
}
cat('</div></div>\n')

cat('</div>\n')
```

### Most Saturated (Vivid) vs. Least Saturated (Monochrome)

```{r}
#| label: vivid-vs-mono
#| results: asis
#| echo: false

vivid_albums <- pop_albums |> arrange(desc(saturation)) |> slice_head(n = 3)
mono_albums <- pop_albums |> arrange(saturation) |> slice_head(n = 3)

cat('<div style="display: grid; grid-template-columns: 1fr 1fr; gap: 24px; margin-bottom: 24px;">\n')

# Vivid column
cat('<div style="background: #fff7ed; border: 1px solid #fed7aa; border-radius: 8px; padding: 16px;">\n')
cat('<h4 style="margin-top:0; color: #ea580c;">🎨 Most Vivid & Colorful</h4>\n')
cat('<div style="display: flex; gap: 12px; flex-wrap: wrap;">\n')
for (i in seq_len(nrow(vivid_albums))) {
  cat(paste0(
    '<div style="text-align: center; width: 110px;">',
    '<img src="', vivid_albums$cover_thumbnail[i], '" style="width: 100px; height: 100px; object-fit: cover; border-radius: 6px; border: 1px solid #f97316;"/>',
    '<div style="font-size: 0.75em; font-weight: bold; margin-top: 4px; overflow: hidden; text-overflow: ellipsis; white-space: nowrap;">', htmltools::htmlEscape(vivid_albums$album[i]), '</div>',
    '<div style="font-size: 0.7em; color: #c2410c;">', vivid_albums$saturation[i], '% Saturation</div>',
    '</div>'
  ))
}
cat('</div></div>\n')

# Monochrome column
cat('<div style="background: #f1f5f9; border: 1px solid #cbd5e1; border-radius: 8px; padding: 16px;">\n')
cat('<h4 style="margin-top:0; color: #475569;">📷 Grayscale / Black & White</h4>\n')
cat('<div style="display: flex; gap: 12px; flex-wrap: wrap;">\n')
for (i in seq_len(nrow(mono_albums))) {
  cat(paste0(
    '<div style="text-align: center; width: 110px;">',
    '<img src="', mono_albums$cover_thumbnail[i], '" style="width: 100px; height: 100px; object-fit: cover; border-radius: 6px; border: 1px solid #94a3b8;"/>',
    '<div style="font-size: 0.75em; font-weight: bold; margin-top: 4px; overflow: hidden; text-overflow: ellipsis; white-space: nowrap;">', htmltools::htmlEscape(mono_albums$album[i]), '</div>',
    '<div style="font-size: 0.7em; color: #64748b;">', mono_albums$saturation[i], '% Saturation</div>',
    '</div>'
  ))
}
cat('</div></div>\n')

cat('</div>\n')
```

---

## 3. Compare by Palette Style

```{r}
#| label: palette-bar
#| echo: false
#| fig-height: 3.5

pop_albums |>
  count(palette, sort = TRUE) |>
  ggplot(aes(x = fct_reorder(palette, n), y = n, fill = palette)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Cool Blue" = "#2563eb",
    "Dark & Moody" = "#1e293b",
    "Green" = "#16a34a",
    "Minimalist White" = "#94a3b8",
    "Monochrome / Gray" = "#64748b",
    "Pink & Magenta" = "#db2777",
    "Warm Red & Orange" = "#ea580c",
    "Warm Yellow & Gold" = "#ca8a04"
  )) +
  labs(
    title = "Distribution of Color Styles in 1,000 Pop Albums",
    x = NULL,
    y = "Number of Album Covers"
  ) +
  theme_minimal(base_size = 11)
```

---

## 4. Search and Filter Table

Use the filter boxes at the top of each column (or the search bar) to group and compare specific styles, artists, or years.

```{r}
#| label: album-table
#| echo: false

pop_albums |>
  mutate(
    Cover = sprintf(
      '<a href="%s" target="_blank"><img src="%s" width="48" height="48" style="border-radius: 4px; object-fit: cover;" loading="lazy" alt="Cover"/></a>',
      cover_full,
      cover_thumbnail
    ),
    Color = sprintf(
      '<span style="display:inline-block; width:22px; height:22px; background-color:%s; border:1px solid #ccc; border-radius:3px;" title="%s"></span>',
      hex_color, hex_color
    )
  ) |>
  select(
    Cover,
    Artist = artist,
    Album = album,
    Year = year,
    Palette = palette,
    Color,
    `Brightness (%)` = brightness,
    `Vibrancy (%)` = saturation
  ) |>
  datatable(
    filter = "top",
    escape = FALSE,
    rownames = FALSE,
    options = list(
      pageLength = 15,
      lengthMenu = c(15, 30, 50, 100),
      autoWidth = TRUE,
      columnDefs = list(
        list(orderable = FALSE, targets = c(0, 5)),
        list(width = "55px", targets = 0),
        list(width = "40px", targets = 5)
      )
    )
  )
```
